---
title: "Assignment 1"
author: "Meghan Duarte"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tsibble)
library(feasts)
```



```{r}
BAFRA <- read.csv('BirthsAndFertilityRatesAnnual.csv')
# check the data with glimpse()
BAFRA |>
  glimpse()
BAFRA
# this code below is not good:
#filtered_data <- Data %>% filter(DataSeries %in% c("Total Fertility Rate (TFR)","Total Live-Births"))
#filtered_data

#rownames(filtered_data)[rownames(filtered_data) == "Total Fertility Rate (TFR)"] <- "TFR"

#filtered_data
#rownames(filtered_data) <- c("TFD", "TLB")
#filtered_data
```

clean the data

- filter the data so we just have TFR and TLB.

```{r}
clean_1 <- BAFRA |>
  filter(DataSeries %in% c("Total Fertility Rate (TFR)", "Total Live-Births")) 
```

- Having each individual year as a row is confusing. Let's kind of turn the data around so that the years are columns. 

```{r}
clean_2 <- pivot_longer(clean_1,
                        cols = starts_with("X"),
                        names_to = "Year",
                        values_to = "Value",
                        values_transform = list(Value = as.character))
clean_2
```

- Let's get rid of the X in each year
```{r}
clean_3 <- mutate(clean_2, Year = str_remove(Year, "X"))
clean_3
```

- Before, some Year values were <dbl> and some were <chr>. I transformed them all the characters to turn Year into a column, let's fix that now. Change Value to <dbl> too while we're at it.

Google says to try:
df <- df %>% mutate(column_name = as.numeric(column_name))


```{r}
clean_4 <- clean_3 |> mutate(Year = as.numeric(Year), Value = as.numeric(Value))
clean_4
```

- Now, give TFR and TLB their own columns

This code below tells R to look at DataSeries column and makes the values in their into their own columns. It uses the values from "Value" to fill up those columns.  
```{r}
clean_data <- clean_4 |>
  pivot_wider(names_from = DataSeries, values_from = Value)
clean_data
```

Okay, now we have a dataset we can work with!

Let's turn it into a tsibble now.

```{r}
singapore <- clean_data |>
  as_tsibble(index = Year)
singapore
```

Let's look at TFR

```{r}
singapore |> autoplot(`Total Fertility Rate (TFR)`)
```

Now let's look at TLB
```{r}
singapore |> autoplot(`Total Live-Births`)
```




